---
title: "Plotting functions in REVOLVER"
author: "Giulio Caravagna"
institute: "Institute for Cancer Research"
date: "June 10, 2018"
email: "giulio.caravagna@ed.ac.uk"
output:
  github_document:
    toc: true
    toc_depth: 2
#output:
#  prettydoc::html_pretty:
#    theme: cayman
#    highlight: github
---


All the examples refer to the breast cancer cohort, available in the package release.


```{r, echo = TRUE, warning=FALSE}
# Load REVOLVER and disable crayon's coloured output that renders badly in HTML,
# as well as REVOLVER's progress bars...
library(revolver)

options(crayon.enabled = FALSE)
options(revolver.progressBar = FALSE)

data("Breast.fit") # Yates et al.
Breast.fit
```
There are three types of plotting functions in REVOLVER:

* functions  ``revolver_report_xxx`` where ``xxx`` is one of ``patient``, ``fit`` or ``clusters``. These functions allow to produce PDFs with almost all the relevant information (sort of multiplots); these functions are handy during to batch PDF creations during cohort analysis' tasks.

* to create PDF reports REVOLVER relies on simpler plotting functions named  ``revolver_plt_xxx``, where ``xxx`` is a keyword that identifies the plot. These functions plot by default to the graphic device, but often accept a  ``file`` parameter to save the plot to PDF; in that case, the device is created with a call to ``pdf`` and usually a ``cex`` parameter is available to control for the size of the output PDF; 

* Because REVOLVER has S3 objects, we have created  ``plot`` functions  that provide easy access to some of the above  ``revolver_plt_xxx`` functions. 

# PDF reports: ``revolver_report_xxx``

To assemble PDFs, REVOLVER uses [PDFjam](https://warwick.ac.uk/fac/sci/statistics/staff/academic-research/firth/software/pdfjam/), which is available for all Unix-like systems, and Windows (via Cygwin); check that you have it installed in your system. 

**Report on data and trees for a patient.**
```{r, results='hide'}
revolver_report_patient(Breast.fit, 'PD9770', cex = 1.5)
```
Example output: ["REVOLVER-report-patient-data-models-PD9770.pdf"](https://github.com/caravagn/revolver.misc/blob/master/vignette_plotting_functions/REVOLVER-report-patient-data-models-PD9770.pdf).

**Report on fit, evolutionary trajectories (exploded) and information transfer for a patient.**
```{r, results='hide'}
revolver_report_fit_patient(Breast.fit, 'PD9770', cex = 3)
```
Example output: ["REVOLVER-report-fit-patient-PD9770.pdf"](https://github.com/caravagn/revolver.misc/blob/master/vignette_plotting_functions/REVOLVER-report-fit-patient-PD9770.pdf).


**Assembling your own PDFs.** You can assemble a bunch of PDFs via REVOLVER's function ``jamPDF``. 

```{r, eval = FALSE}
# Take file "1.pdf", "2.pdf", merge them to a unique PDF with layout "2x1"
jamPDF(
  in.files = c("1.pdf", "2.pdf"),
  out.file = 'newfile.pdf',
  layout = '2x1',                 # the layout (matrix style)
  delete.original = TRUE          # delete original files
  crop.white = TRUE,              # crop white borders for overly large PDF canvases
  page = 'a4'                     # resize each page to fit A4 paper
)                 
```


# Plotting functions ``revolver_plt_xxx``

### Data and trees for a patient

**Plotting a REVOLVER tree.** The top-ranked tree for patient ``PD14753``, among all the ones available is

```{r, fig.width=5, fig.height=5}
tree = Breast.fit$phylogenies$PD14753[[1]]

revolver_plt_tree(x = tree, type = 'binary')
```

A tree is a built over the groups annotated in the data, here named 1, 2, 3, ..., 11. Each node is coloured only if, whithin the gruop, there are annotated drivers. Here all groups but 4, 7, 8 and 9 have a driver annotated. Edges are coloured according to the trajectories that we can extract from this tree. For instance, the trajectories starting from 3 and finishing in 11, is coloured orange to highlight that 3 is the upstream group; GL is a special node that means germline.

The plot can also report violations that the tree has with respect to the data: ``type = 'binary'`` tells which type of data we are using, and defines how to compute violations which are normally returned by calling ``stats.rev_phylo``. For ``binary`` data, which is the on used for breast cancer, the field reported is ``$Suppes``, and a violation is when the inequality holds as ``<``. For  ``CCF``, the field reported is ``$CCF.pigeonhole`` and violations are due to the pigeonhole principle.
```{r}
stats.rev_phylo(tree)
```

**Plotting data for a patient.** You can plot a ``pheatmap`` of the input data, annotated with driver and clonality status. In that plot each row is a group of alterations that occur in the same set of samples (with binary data), or a group of alterations associated to the same cluster/ clone (with CCF data). Every group may have or not drivers associated; columns represent samples (regions, for instance). 

```{r, results='hide'}
revolver_plt_patient_data(Breast.fit, patient = 'PD14753')
```

**Plotting all the trees of a patient (the distribution).** We  plot the score of the trees for a patient with a barplot (sorted). Each bar is a tree, and is coloured according to its information transfer; repeated colours mean that more than one tree transfer the same orderings among the drivers. The more colours then, the more combinations of transfers are available. 


```{r,  fig.width=6, fig.height=3, results='hide'}
# Here we have 8trees with 3 different information transfers
revolver_plt_patient_trees_scores(Breast.fit, patient = 'PD14768')
```

We can visualize the top ``K`` trees for a patient by iteratively applying tree plotting functions, as wrapped via ``revolver_plt_patient_trees``. The number of plots is limited above by ``K``, and the number of trees available for the patient. 

```{r, fig.width=15, fig.height=15, results='hide'}
# Inspect the information transfer (observe the different ordering of the tree nodes).
revolver_plt_patient_trees(Breast.fit, patient = 'PD14768', max.phylogenies = 12, file = NA, cex = 1)
```

So, ``PD14768`` has some combinations of transfer repeated because group 2 has no drivers, and thus all the models that predict a branch towards 3 and 4, leads to the same trasnfer. 

### Visualizing fits


**Plotting a patient's fit, its trajectories and the information transfer.** The fit of a patient is accessible via a direct function, which wraps a call to ``revolver_plt_tree`` 
```{r, fig.width=5, fig.height=5}
 revolver_plt_fit_patient(Breast.fit, patient = 'PD9850')
```

From the model fit, REVOLVER  compute the expansion of all the nodes of the tree with at least  one  drivers associated, which leads us to the evolutionary trajectories of the model. In this case, we "expand" the coloured  nodes, and project inside each expanded node the possible trajectories that could have lead to the accumulation of the annotated drivers. When there is no unique consensus on that, the expanded structure can be a general graph as in the case of the red node. In this case the only evidence that REVOLVER finds (from other patients) is that ``ARID1B`` is a driver that triggers after ``TP53``; for the others we have uncertainty or no evidence, and hence they are connected to the germline node, as well as to the downstream node. For group 2, blue, we can instead retrieve a linear ordering for the drivers annotated in the group.

```{r, fig.width=6, fig.height=7}
revolver_plt_trajectories_patient(Breast.fit, patient = 'PD9850')
```

At the same time, we can extract the information transfer from  the model fit, which consists in dropping the nodes from the above plot when they do not harbour any driver. The plot is annnotated: for each driver we report the number of times it is clonal and subclonal, and for each edge its frequency across the whole cohort (i.e., how often that is found in other patients). So for instance the event annotated as ``TP53`` is clonal in 27 patients, subclonal in 3, and every time that it is clonal it is connected to germline (giving support for the hypothesis of that event triggering tumour initiation).


```{r, fig.width=6, fig.height=7}
revolver_plt_itransfer_patient(Breast.fit, patient = 'PD9850')
```




**Plotting w, the penalty of the fit.** The penalty-per-patient is shown with a simple barplot computed by marginalizing **w** over the alterations 
```{r, fig.width=3, fig.height=6}
revolver_plt_penalty_barplot(Breast.fit)
```

While the penalty across patients and alterations is shown with a ``pheatmap``. This can be plot for the models fit at the end of the TL fit (likelihood fit + expansion step), by default, or  you can use ``type="before.expansion"`` to use the models computed after likelihood fit (without the expansion). If you prefer, you can  plot the normalized version of  **w** with ``normalized = TRUE``.

```{r, fig.width=9, fig.height=9}
revolver_plt_penalty_matrix(Breast.fit)
```

**Plotting the index of Divergent Evolutionary Trajectories (DET).** The DET for the  cohort  is estimated via a non-parametric bootstrap with ``N`` resamples, and visualized in a histogram.

```{r, fig.width=6, fig.height=4}
revolver_plt_DET_index(Breast.fit, N = 100)
```

For each driver, it is reported as estimated empirically from data

```{r, fig.width=3, fig.height=6}
revolver_plt_DET_index_driver(Breast.fit)
```

  <!-- revolver_penaltyPlot(fit) -->
  <!-- plot(fit, out.file = paste(cohort.name, '.fit.pdf', sep = ''), plot.stat = TRUE, layout = '1x1',  palette = 'Dark2') -->
  
  
  
  

### Visualizing clusters

**Plotting features from REVOLVER clustering.** You can visualize REVOLVER's clusters and trajectories (features) to spot relevant trajectories in each group. This is probably one of the most important plots that you will produce from your analysis.

```{r, fig.width=20, fig.height=20}
# We suppress the warnings just because we use some UNICODE chars to plot edges, and this give rise to a long list of useless warnings due to R's output graphics.
suppressWarnings(
  revolver_plt_rclusters(Breast.fit, cutoff.features_annotation = 3, cex = 1)
)
```

**Plotting dendograms for REVOLVER and occurence-based clusters.** You can plot dendrograms for REVOLVER's clusters, or for clusters computed by scanning the patterns of occurence of the annotated drivers (binary, or clonal/ subclonal), coloured with REVOLVER clustering assignemnt. We usually use these to see how much difference it makes to consider the evolutionary trajectories, rather than the patterns of occurrences of the annotated drivers.


```{r}
revolver_plt_dendogram(Breast.fit, type = 'REVOLVER')
```

The other types of clusterings are accessible with the ``type`` parameter.
```{r, fig.width=12, fig.height=3}
par(mfrow = c(1,2))
revolver_plt_dendogram(Breast.fit, type = 'binary')
revolver_plt_dendogram(Breast.fit, type = 'clonality')
par(mfrow = c(1,1))
```

**Bannerplot for REVOLVER clusters.** This is a classical bannerplot that is sometimes used to determine the number of clusters in the data (especially with static cut methods).
```{r, fig.width=5, fig.height=5}
revolver_plt_rbannerplot(Breast.fit, cex = 1)
```

**Tanglegrams to compare REVOLVER and occurence-based clusters.**  Tanglegrams against clusters obtained by 
patterns of occurrences of the input alterations highlight how groups of patients change when you use REVOLVER. Use ``type = 'clonality'`` to compare against clonal/ subclonal patterns.

```{r}
revolver_plt_compare_dendograms(Breast.fit, cex = 1, type = 'binary')
```



**Plotting REVOLVER's evolutionary distance.** 
You can display the  distance that has been used to compute the clusters, for each pair of patients. The parameters used to compute the distance are reported on top of the plot.

```{r}
revolver_plt_evodistance(Breast.fit, cex = 1)
```


**Plotting the models trajectories present in each of REVOLVER's groups.** It might be convenient to plot the fit of each patient, as separate files divided by REVOLVER's cluster. This helps spot the features that define the clusters. We have prepared a function that makes this plot automatically; the function will also plot a table reporting the  for the most frequent trajectories. As output, the function creates a set of `K` PDFs -- if there are `K` clusters identified by REVOLVER -- all named with a suffix that you specify by the string ``file.suffix``.  We do not show execution of this function here.

```{r, fig.width=15, fig.height=15, eval = FALSE}
revolver_plt_fit_by_group(Breast.fit, cex = 1, file.suffix = "REVOLVER-Clusters-FitsPerCluster.pdf")
```


### Visualizing confidence (jackknife)

**Plotting co-clustering statistics.** The co-clustering statistics -- i.e., the probability or clustering together two samples, across each resample --  with all entries above 0.7 explcitly annotated as  patient-by-patient matrix 
```{r, fig.width=9, fig.height=9}
revolver_plt_jackknife_coclust(Breast.fit, cutoff.annotate.numbers = 0.7)
```

From the above numbers, we can compute a boxplot per cluster which gives us a mean/ median stability value per cluster. This seems a sensitive summary statistics to report for each group.
```{r, fig.width=6, fig.height=3}
revolver_plt_jackknife_coclust_bplot(Breast.fit)
```

**Plotting edge statistics.** Concerning edges, the probability of detecting each edge across resamples is a driver-by-driver matrix

```{r, fig.width=11, fig.height=9}
revolver_plt_jackknife_edge_prb(Breast.fit)
```

while the jackknife statistics for the number of patients with an edge is a barplot where we  annotate all edges with mean count above ``cutoff.annotate.numEdges``

```{r, fig.width=5, fig.height=7}
revolver_plt_jackknife_edge_counts(Breast.fit, cutoff.annotate.numEdges = 4)
```

## Plot functions for S3 objects

* For class ``rev_phylo``, function ``plot`` will use ``revolver_plt_tree`` to plot the tree.
```{r, eval = FALSE}
plot(Breast.fit$phylogenies$PD9193[[1]])
```

* For class ``rev_cohort``, function ``plot`` will automatize the creation of PDF reports for a set of patients, via repeated calls to ``revolver_report_patient``. This function outputs only to file.

```{r, results='hide', eval = FALSE}
#  To show this we convert ``Breast.fit`` to the proper class
conv = Breast.fit
class(conv) = 'rev_cohort'

plot(conv, patients = Breast.fit$patients)
```
  

* For class ``rev_cohort_fit``, function ``plot`` will automatize the creation of a PDF report for a set of patients, via repeated calls to ``revolver_report_fit_patient``.  This function outputs only to file.

```{r, results='hide', eval = FALSE}
class(Breast.fit)
plot(Breast.fit, patients = Breast.fit$patients)
```
  